---
title: "A Study on Flocking behavior and Sexual Dimorphism in Taiwanese Birds"
author: "Ming Pei Li"
date: "2024-12-01"
output: html_document
---

Introduction
I hypothesize that bird flocking behavior differs between urban and natural habitats, primarily due to variations in predation risk and food resources. (Additional literature will be provided to elaborate on 



Methods

Steps for R code

Load the necessary packages.
```{r}
library (dplyr)
library (ggplot2)
```

```{r}
bird_2 <- read.table ("D:/碩士資料庫/碩三上課程/Ocean R/Ocean-R-Final-project/dwca-trait_454-v1.68/measurementorfacts.txt", header = TRUE, sep = "\t", fill = TRUE)

bird_3 <-  read.table ("D:/碩士資料庫/碩三上課程/Ocean R/Ocean-R-Final-project/dwca-trait_454-v1.68/taxon.txt", header = TRUE, sep = "\t", quote = "", fill = TRUE)
```

Filter the bird data for conspecific flocking behavior during the non-breeding season.
```{r}
co_flocks_nbd <- bird_2 %>% 
  filter (measurementType %in% c("Flocks_ConspecificsNonBreedingSeason"))
```

Filter the bird data for non-territorial birds during the non-breeding season.
```{r}
nt_nbs <- bird_2 %>% 
  filter (measurementType == "Territoriality_NonBreedingSeason" & measurementValue == 0)
```

Classify birds living in anthropogenic habitats (such as farmlands, urban parks, etc.) as "Anthropogenic habitat." Classify birds not living in these anthropogenic habitats as "Natural habitat."
```{r}
human_habitat <- bird_2 %>% 
  filter (measurementType %in% 
            c("Habitat_DrylandFarming",
              "Habitat_UrbanParks",
              "Habitat_UrbanRuralArea",
              "Habitat_AquaculturePonds",
              "Habitat_SaltExploitationSites",
              "Habitat_PaddyField") & 
            measurementValue == 1) %>% 
  mutate (habitat = "Anthropogenic habitat")

nonhuman_habitat <- bird_2 %>% 
  filter (grepl ("Habitat", measurementType) 
          & !(measurementType %in% c("Habitat_DrylandFarming",
                                     "Habitat_UrbanParks",                                 
                                     "Habitat_UrbanRuralArea", 
                                     "Habitat_AquaculturePonds", 
                                     "Habitat_SaltExploitationSites",
                                     "Habitat_PaddyField")) & 
            measurementValue == 1) %>% 
  mutate (habitat = "Natural habitat")
```


Combine all habitat data and determine the primary habitat (either natural or anthropogenic) for each bird.
```{r}
maj_habitat <- 
  rbind (nonhuman_habitat, human_habitat) %>% 
  group_by (id, habitat) %>%
  tally () %>% 
  group_by (id) %>%
  filter (n == max (n) & n != min(n)) %>%
  select (id, majority_habitat = habitat)
```

Merge the flocking behavior data with the primary habitat data and filter for non-territorial birds during the non-breeding season.
```{r}
co_flock_habitat <- co_flocks_nbd %>%
  inner_join (maj_habitat, by = "id") %>% 
  inner_join(bird_3, by = "id") %>% 
  filter (id %in% nt_nbs$id)
```

Use the Chi-squared test to compare whether birds from different habitats (natural vs. anthropogenic) show significant differences in flocking behavior.
```{r}
chisq.test (co_flock_habitat$measurementValue, co_flock_habitat$majority_habitat)
```

Use Fisher's Exact Test to compare whether birds from different habitats (natural vs. anthropogenic) show significant differences in flocking behavior, applicable for small sample sizes.
```{r}
fisher.test (co_flock_habitat$measurementValue, co_flock_habitat$majority_habitat)
```

Use a Generalized Linear Model (GLM) with bird flocking behavior (yes/no) as the dependent variable and primary habitat (natural or anthropogenic) as the independent variable to further examine the impact of habitat on flocking behavior.
```{r}
bird_glm <- glm (measurementValue ~ majority_habitat,
                 data = co_flock_habitat, 
                 family = binomial)
summary (bird_glm)
```

Calculate the frequency of combinations between different habitats and whether flocking behavior (yes/no) is present.
```{r}
stacked_data <- co_flock_habitat %>%
  group_by (majority_habitat, measurementValue) %>%
  summarise (count = n ()) %>%
  ungroup ()
```

Calculate the proportion of birds exhibiting flocking behavior in different habitats.
```{r}
stacked_data <- stacked_data %>%
  group_by (majority_habitat) %>%
  mutate (percentage = count / sum (count))
```

Plot the proportion of birds exhibiting flocking behavior by habitat type using a stacked bar chart.
```{r}
ggplot(stacked_data, aes(x = majority_habitat, y = percentage, fill = as.factor (measurementValue))) +
  geom_bar(stat = "identity", position = "stack", width = 0.6) +
  scale_fill_manual(values = c("1" = "skyblue", "0" = "lightcoral"),
                    labels = c("1 (Flocking)", "0 (No Flocking)"),
                    name = "Flocking Behavior") +
  labs(title = "Flocking Behavior by Habitat",
       x = "Habitat Type",
       y = "Proportion") +
  scale_y_continuous (limits = c(0, 1.1), breaks = seq (0, 1, 0.2), labels = c("0.00", "0.20", "0.40", "0.60", "0.80", "1.00")) +
  theme_minimal() +
  theme (legend.position = "top",
         panel.grid.minor.y = element_blank (),
         panel.grid.major.x = element_blank (),
         plot.title = element_text (hjust = 0.5, size = 16),
         plot.margin = margin (20, 20, 20, 20),
         axis.title.x = element_text (margin = margin(t = 15), size = 12),
         axis.title.y = element_text (margin = margin(r = 10), size = 12),
         axis.text.x = element_text (size = 12),
         axis.text.y = element_text (size = 12)) +
  annotate ("text", x = 1.5, y = 1.1, label = "*", size = 6, color = "black") +
  annotate ("segment", x = 1, xend = 2, y = 1.05, yend = 1.05, color = "black", linewidth = 0.5) +
  annotate ("text", x = 2, y = 0.1, label = "n = 189", size = 4, color = "black") +
  annotate ("text", x = 1, y = 0.1, label = "n = 20", size = 4, color = "black")
```

Figure 1. The Impact of Habitat on Bird Flocking Behavior
This figure presents the proportion of birds exhibiting flocking behavior during the non-breeding season across different habitats, classified as either anthropogenic habitat (e.g., farmland, urban parks) or natural habitat. The data are shown as a stacked bar plot, where the colors of the bars represent whether flocking behavior was observed (red for flocking, blue for no flocking). The sample sizes for each habitat are indicated at the bottom, with 20 birds in anthropogenic habitats and 189 birds in natural habitats. Asterisks above the bars indicate a statistically significant difference.

Results
Figure 1 shows a significant difference in flocking behavior between birds from different habitats, with a higher proportion of birds exhibiting flocking behavior in natural habitats compared to anthropogenic habitats. The Chi-squared test revealed a significant difference in flocking behavior between habitats (χ² = 5.63, df = 1, p = 0.017). Additionally, Fisher's Exact Test confirmed this result, showing a significant difference in flocking behavior between habitats (p = 0.007, 95% CI: 0.0186–0.743). The Generalized Linear Model (GLM) analysis further supported this finding, indicating that birds in natural habitats have a significantly higher probability of flocking behavior compared to those in anthropogenic habitats (estimate = -1.78, p = 0.019).

Disscussion


References
